Stop Lying to Me: New Visual Tools to Choose the Most Honest Nonlinear Dimension Reduction

### 2025-06-26 paper.qmd
discussion - PH

Two examples illustrating usage are provided: the PBMC3k data where the NLDR is summarizing clustering in \pD{} and hand-written digits illustrating how NLDR represents an intrinsically lower dimensional nonlinear manifold. We demonstrated that the published layout of the PBMC3k is inaccurate, because it grossly exaggerates separation between clusters, and even suggests separation when there is none. This is common when layouts are chosen subjectively -- often a preference for the "prettiest". Our approach provides a way to objectively choose the layout and hopefully avoids the use of misleading layouts in the future. In the hand-written digits we illustrate how our model fit statistics show that a flat disc layout is superior to the curved shaped layouts, and how to identify oddly written "1"s using the residuals of the fitted model.

@@

Most NLDR methods only provide a \gD{} but if a \kD{} ($k>2$) layout is provided the approach developed here could be extended. Binning into cubes could be done in $3\text{-}D$ or higher, relatively easily, and used as a the basis for a wireframe of the fitted model. @barber1996 and the software @stephane2023 have algorithms for convex hulls, which \pD{} which serve as an inspiration. A simpler approach using $k$-means clustering to provide centroids could also be possible, but the complication would be to determine how to connect the centroids into an appropriate wireframe.

>>>>

Two examples illustrating usage are provided: the PBMC3k data where the NLDR is summarizing clustering in \pD{} and hand-written digits illustrating how NLDR represents an intrinsically lower dimensional nonlinear manifold. We examined a typical published usage of UMAP with the PBMC3k dataset (Chen, 2024). As is typical of UMAP layout with default settings, the separation between clusters is grossly exaggerated. The layout even suggests separation where there is none. <!--This is common when layouts are chosen subjectively -- often a preference for the "prettiest".--> Our approach provides a way to objectively choose the layout and hopefully avoids the use of misleading layouts in the future. In the hand-written digits we illustrate how our model fit statistics show that a flat disc layout is superior to the curved shaped layouts, and how to identify oddly written "1"s using the residuals of the fitted model.

@@

Most NLDR methods only provide a \gD{} but if a \kD{} ($k>2$) layout is provided the approach developed here could be extended. Binning into cubes could be done in $3\text{-}D$ or higher, relatively easily, and used as a the basis for a wireframe of the fitted model. @barber1996 (and the associated software @stephane2023) has an algorithm for a convex hull in \pD{}m which serves as an inspiration. A simpler approach using $k$-means clustering to provide centroids could also be possible, but the complication would be to determine how to connect the centroids into an appropriate wireframe.
